---
title: "Microarray Data - Il1b treated stomach stroma - DGE Analysis with LIMMA"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, messages=FALSE}
rm(list=ls())
library(limma)
library(readxl)
library(pheatmap)
library(xtable)
library(reshape2)
library(knitr)
library(writexl)
library(EnhancedVolcano)

load("../Data/Processed/Il1b_stroma_preprocessed_data.Rdata")

result_folder = "../Results"
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)
```

# Introduction

This is data from mouse experiments hybridized as dual channel libraries to Agilent 074809 (SurePrint G3 Mouse GE v2 8x60K) micro arrays. 


# Diffential gene expression

Standard dual-channel analysis of differential expression data using LIMMA is applied for comparisons. 

We do the following comparisons: 

  - Il1b treated vs. untreated stroma

```{r}
all_results = list()

controls = MA$genes$ControlType!=0
MA.avg <- avereps(MA[!controls,], ID=MA[!controls,]$genes$ProbeName)
```


```{r}
#######################################################################
sel_samples = rownames(ed)
ed_tmp = ed[sel_samples,]

design = ifelse(ed_tmp$Cy5_condition=="IL1b", 1, -1)
biolrep <- apply(ed_tmp, 1, function(x) {sort(c(x["Cy5_replicate"], x["Cy3_replicate"])) |> paste0(collapse="_")}) |> factor()

corfit <- duplicateCorrelation(MA.avg, design, ndups = 1, block = biolrep)
fit <- lmFit(MA.avg, design, block = biolrep, cor = corfit$consensus)
#fit <- lmFit(MA.avg[, sel_samples], design)
fit <- eBayes(fit)
res = topTable(fit, adjust = "BH",number=nrow(fit))

all_results[["Il1b_vs_Ctrl"]] = res
```


## Volcano plots

```{r, volcano, echo=FALSE, fig.width=8, fig.height=7}
all_target_conditions = names(all_results)
#par(mfrow=c(1,3), mar=c(8,4,4,2))

for (tc in all_target_conditions) {
  r = all_results[[tc]]

  p = EnhancedVolcano(r, lab=r$GeneSymbol, x="logFC", y="adj.P.Val", pCutoff = 0.25, ylim = c(NA, 1.5), subtitle = "FDR < 25%")
  print(p)
}
```


```{r, fig.width=6, fig.height=7}
sel_targets <- c("PTGS2
PTGES2
PTGES
") |> strsplit("\n") |> unlist() |> unique() |> (\(x) paste0(substr(x,1,1), tolower(substr(x,2,nchar(x))) ))()

r = all_results$Il1b_vs_Ctrl
p = EnhancedVolcano(r, lab=r$GeneSymbol, x="logFC", y="adj.P.Val", pCutoff = 0.2, selectLab = sel_targets, 
                    ylim = c(NA, 1.5), xlim = c(-4.5, 4.5), 
                    subtitle = "FDR < 25%", FCcutoff = 0.5, 
                    drawConnectors = TRUE, widthConnectors = 1.0, colConnectors = 'black',arrowheads = FALSE,
                    boxedLabels = TRUE)
print(p)

pdf(file.path(result_folder, "Volcano_PTGS2.pdf"), width=6, height = 7)
print(p)
dev.off()

# 
# p = EnhancedVolcano(r, lab=r$GeneSymbol, x="logFC", y="P.Value", pCutoff = 0.001, selectLab = sel_targets, drawConnectors = T, subtitle = "FDR < 25%", FCcutoff = 0.5)
# print(p)

```



```{r, DE_combined, echo=FALSE, results="hide"}
###############################################################################################################
# Write Result Files
###############################################################################################################

all_DE_results_tmp = list()
for (tc in all_target_conditions) {
  tmp = all_results[[tc]]
  tmp$condition = tc
  all_DE_results_tmp[[tc]] = tmp
}
all_DE_results_ts = do.call(rbind, all_DE_results_tmp)
all_DE_results_ts$DE_class = ifelse(all_DE_results_ts$adj.P.Val>0.01, "not sign", ifelse(all_DE_results_ts$logFC>0,ifelse(all_DE_results_ts$logFC>1,"Up","Up_weak"), ifelse(all_DE_results_ts$logFC < -1,"Down", "Down_weak")))
#all_DE_results_ts$DE_class = ifelse(all_DE_results_ts$adj.P.Val>0.01, "n.s.", ifelse(all_DE_results_ts$logFC > 0,"Up","Down"))
#agg_fun = function(x) paste(unique(x),collapse=";")
agg_fun = function(x) ifelse("Down" %in% x, "Down",ifelse("Up" %in% x, "Up","not_significant."))
#agg_fun = function(x) paste(unique(sort(x)), collapse=",")
all_DE_results_sw = dcast(all_DE_results_ts, GeneSymbol ~ condition, value.var="DE_class", fun.aggregate=agg_fun)
```


```{r, echo=FALSE}
R.avg <- avereps(RG$R, ID=RG$genes$ProbeName)
G.avg <- avereps(RG$G, ID=RG$genes$ProbeName)
colnames(R.avg) = paste(ed[colnames(R.avg),]$Cy5,"_Cy5_label",sep="")
colnames(G.avg) = paste(ed[colnames(G.avg),]$Cy3,"_Cy3_label",sep="")

intensity_matrix = cbind(R.avg, G.avg)
norm_intensity_matrix = normalizeBetweenArrays(intensity_matrix, method="quantile")
```


```{r, write_tabs, echo=FALSE}


output_file_prefix = paste(result_folder,"Differential_expression_results_", sep="/")
selected_cols = c("ProbeName", "GeneSymbol", "GeneDescription","EntrezID","logFC","AveExpr","t","P.Value","adj.P.Val","GencodeM22MappingStatus","GeneSymbol_Gencode","Transcript_Type" )
for (tc in all_target_conditions) {
  write.table(all_results[[tc]][,selected_cols], file=paste(output_file_prefix, tc, ".txt", sep="" ), row.names=F , sep="\t", dec=".")
}

short_names = list()
for (i in names(all_results)) {
  ii = gsub(" ","_", gsub(",","_",i))
  short_names[[i]] = ii
}

DGE_Excel_filename = file.path(result_folder,"DiffExpression.xlsx")
# txt2xlsx_command = paste("python /home/hilmar/Work/Python/tools/txt2xlsx.py -o",DGE_Excel_filename," -L", paste(unlist(short_names),collapse=","),  paste(paste('"', output_file_prefix, names(short_names), '.txt"', sep="" ), collapse=" "), sep=" ")
# system(txt2xlsx_command)

write_xlsx(all_results, path=DGE_Excel_filename)


##############################################
# Normalized expression values per sample
anno = unique(RG$genes[,!colnames(RG$genes) %in% c("Row","Col","Start","index")])
rownames(anno) = anno$ProbeName
norm_exp_mat = data.frame(ProbeName = rownames(norm_intensity_matrix), log2(norm_intensity_matrix), row.names=NULL, stringsAsFactors = F)
norm_exp_mat$GeneSymbol = anno[as.character(norm_exp_mat$ProbeName), "GeneSymbol"]
filename = paste(result_folder,"Normalized_expression_data.txt",sep="/")
write.table(norm_exp_mat, file=filename,sep="\t",row.names=F)

```

```{r}
output_file = paste(result_folder,"DGE_analysis_image.Rdata", sep="/")
save(all_results, RG, MA, ed, file=output_file)

filename = paste(result_folder,"DGE_results.Rdata",sep="/")
exp_design = ed
save(all_results, exp_design, norm_intensity_matrix, MA, file=filename)
```


# Session Info

```{r}
sessionInfo()
```