---
title: "Microarray Data of Il1b on stromal cells - GSEA analysis on DGE results - Plots"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, warning=FALSE, message=FALSE}
rm(list=ls())

#suppressMessages(library(xlsx))
library(ggplot2)
library(knitr)
library(scales)
library(fgsea)
library(data.table)
library(gridExtra)
library(homologene)

#source("./fGSEA_plot_functions.R")
source("../../r_utils/fGSEA_plot_functions.R")

result_folder = "../Results"
if (!file.exists(result_folder)) dir.create(result_folder, recursive=T)

set.seed(12345)
```

# Overview

This is data from mouse experiments hybridized as dual channel libraries to Agilent 074809 (SurePrint G3 Mouse GE v2 8x60K) micro arrays. 
 

For analysis, a preranked analysis using the R package fGSEA was used that should give similar results to preranked analysis in standard GSEA. 


```{r}
load("../Results/DGE_results.Rdata")
```


```{r}
output_folder = "../Results"
GSEA_datafile = file.path(output_folder,"fGSEA_Results.Rdata")
load(GSEA_datafile)
load(file.path(output_folder, "Geneset_collections.Rdata"))
```



# MSigDB gene sets

Since MSigDB gene sets use human gene symbols to map genes to pathways we translated mouse symbols to homologous human symbols using HomologeneDB from NCBI (build 68). 



# Plots for selected gene sets

```{r}
sel_genesets_1 = "HALLMARK_INFLAMMATORY_RESPONSE
HALLMARK_TNFA_SIGNALING_VIA_NFKB
HALLMARK_INTERFERON_ALPHA_RESPONSE
HALLMARK_INTERFERON_GAMMA_RESPONSE
HALLMARK_IL6_JAK_STAT3_SIGNALING
HALLMARK_IL2_STAT5_SIGNALING
HALLMARK_TGF_BETA_SIGNALING
HALLMARK_WNT_BETA_CATENIN_SIGNALING
HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION
VECCHI_GASTRIC_CANCER_ADVANCED_VS_EARLY_DN
VECCHI_GASTRIC_CANCER_ADVANCED_VS_EARLY_UP
VECCHI_GASTRIC_CANCER_EARLY_DN
VECCHI_GASTRIC_CANCER_EARLY_UP
LEE_BMP2_TARGETS_DN
LEE_BMP2_TARGETS_UP
HALLMARK_FATTY_ACID_METABOLISM
HALLMARK_OXIDATIVE_PHOSPHORYLATION
"
selected_genesets =  c(unlist(strsplit(sel_genesets_1, "\n")) , names(gset_collections_mouse[["CUSTOM_MOUSE"]]), names(gset_collections_human[["CUSTOM_HUMAN"]]) )

#ct <- names(gset_collections_human$CELLTYPES)
#selected_genesets <- append(selected_genesets, ct[grepl("BUSSLINGER_GASTRIC_",ct)])
```

```{r, fig.width=18, fig.height=6, results="hide"}
all_plots = list()
for (sel_comparison in names(all_results)) {
  tmp = all_results[[sel_comparison]]
  # tmp2 = merge(tmp, mouse_to_human, by="GeneSymbol")
  # tmp2 = subset(tmp2, !is.na(symbol_1))
  ranks_mouse = unlist(tapply(tmp$t, tmp$GeneSymbol, function(x) x[which(abs(x)==max(abs(x)))] ))

  m2h <- mouse2human(tmp$GeneSymbol, db = homologeneData2)

  tmp2 = merge(tmp, m2h, by.x="GeneSymbol", by.y="mouseGene")
  tmp2 = subset(tmp2, !is.na(humanGene))
  ranks_human = unlist(tapply(tmp2$t, tmp2$humanGene, function(x) x[which(abs(x)==max(abs(x)))] ))

  gg <- strsplit(sel_comparison, "_vs_") |> unlist()
  up_group <- gg[1]
  down_group <- gg[2]
  
  for (g in selected_genesets) {
    rr = subset(all_gsea_results[[sel_comparison]], pathway==g)
    loc = unlist(lapply(gset_collections_human, function(x) which(g == names(x))))
    if(length(loc)>0) {
      sel_genes = unlist(gset_collections_human[[names(loc)]][loc])
      ranks = ranks_human
    } else {
      loc = unlist(lapply(gset_collections_mouse, function(x) which(g == names(x)) ))
      sel_genes = unlist(gset_collections_mouse[[names(loc)]][loc])
      ranks = ranks_mouse

    }
    p = myPlotEnrichment(sel_genes, ranks, group_pos = up_group, group_neg = down_group, color_gene_bars = F, color_pos = "black", color_neg = "black")
    p1 = p + ggtitle(paste0(g, "\n", sel_comparison), paste("FDR=",prettyNum(rr$FDR_global, digits=2),", p=",prettyNum(rr$pval, digits=2), ", ES=",prettyNum(rr$ES, digits=2), ", NES=",prettyNum(rr$NES, digits=2) ) )  + theme(title = element_text(size=12))
    #print(p1)
    all_plots[[paste0(g,"_",sel_comparison)]] = p1
  }  
}

all_plots = all_plots[order(names(all_plots))]
print(marrangeGrob(all_plots, ncol=3, nrow=1))

```


### PDF export

All plots have been exported to corresponding files in PDF format in folder `r result_folder`.

<!-- ```{r} -->
<!-- for (g in selected_genesets) { -->
<!--   rr = subset(all_gsea_results[[sel_comparison]], pathway==g) -->
<!--   loc = unlist(lapply(gset_collections_human, function(x) grep(g, names(x)))) -->
<!--   sel_genes = unlist(gset_collections_human[[names(loc)]][loc]) -->
<!--   p = myPlotEnrichment(sel_genes, ranks, linewidth = 2) -->
<!--   pdf(file=file.path(result_folder, paste0(sel_comparison,"_",g, ".pdf")), width=8, height = 6) -->
<!--   print(p + ggtitle(g, paste("FDR=",prettyNum(rr$FDR_global, digits=2),", p=",prettyNum(rr$pval, digits=2), ", ES=",prettyNum(rr$ES, digits=2), ", NES=",prettyNum(rr$NES, digits=2) ) ) ) -->
<!--   #print(p + ggtitle(g, paste("Adj.pval=",prettyNum(rr$padj, digits=2), "ES=",prettyNum(rr$ES, digits=2), ", NES=",prettyNum(rr$NES, digits=2) ) ) ) -->
<!--   dev.off() -->
<!-- } -->
<!-- ``` -->


```{r}
for (g in names(all_plots)) {
  pdf(file=file.path(result_folder, paste0(make.names(g), ".pdf")), width=8, height = 6)
  print(all_plots[[g]])
  dev.off()
}
```


# Software versions

```{r}
sessionInfo()
```
