---
title: "Microarray Data - Il1b treated stomach stroma - analysis with LIMMA"
author: "Hilmar Berger"
output: 
 html_document:
    toc: true
    toc_depth: 3
    toc_float: false
    number_sections: true
    code_folding: hide
    
pdf_document:
    fig_caption: true
date: "`r format(Sys.time(), '%d %B, %Y %H:%M:%S')`"
---

```{r, messages=FALSE}
rm(list=ls())
library(limma)
library(readxl)
library(pheatmap)
library(xtable)
library(reshape2)
library(knitr)
```

# Introduction

This is data from mouse experiments hybridized as dual channel libraries to Agilent 074809 (SurePrint G3 Mouse GE v2 8x60K) micro arrays. 

```{r, messages=FALSE}
# set up the experimental design
exp_design = as.data.frame(read_excel("../Data/Raw/samples.xlsx",sheet = 1), stringsAsFactors=F)
colnames(exp_design) = make.names(colnames(exp_design))
rownames(exp_design)=gsub("\\.txt", "", as.character(exp_design$FileName))
#exp_design$FileName = paste(exp_design$Scan.Name, ".txt", sep="")

tmp_cy3 <- strsplit(exp_design$Cy3," ")
tmp_cy5 <- strsplit(exp_design$Cy5," ")

exp_design$Cy3_replicate <- sapply(tmp_cy3, `[`,1)
exp_design$Cy5_replicate <- sapply(tmp_cy5, `[`,1)
exp_design$Cy3_condition <- sapply(tmp_cy3, `[`,2)
exp_design$Cy5_condition <- sapply(tmp_cy5, `[`,2)

exp_design$dye_swap = exp_design$DyeSwap

exp_design$Cy3_label = with(exp_design, paste(Cy3_condition, Cy3_replicate, sep="_"))
exp_design$Cy5_label = with(exp_design, paste(Cy5_condition, Cy5_replicate, sep="_"))

exp_design$array_label = with(exp_design, paste(ifelse(dye_swap, Cy3_label, Cy5_label), "_vs_",ifelse(dye_swap,Cy5_label, Cy3_label),"_",ifelse(dye_swap,"dye_swp","no_swp") ,sep=""))

# read data from raw files
setwd("../Data/Raw/")

RG =read.maimages(exp_design$FileName,source="agilent.median", columns=list(G="gProcessedSignal",R="rProcessedSignal"))
setwd("../../Code/")
ed = exp_design

# fix outdated chip annotations
new_anno_file = "../Data/Raw/074809_60k_annotations_2020-09-18.Rdata"
load(new_anno_file)
old_anno = RG$genes
take_over_cols = colnames(old_anno)[!colnames(old_anno) %in% c("GeneName","Description","SystematicName")]
tmp = old_anno[,take_over_cols]
tmp$index=1:nrow(tmp)
tmp = merge(tmp, anno_tab_final, by.x="ProbeName", by.y="ProbeName", all.x=T, sort=F)
new_col_order = c(take_over_cols, colnames(tmp)[!colnames(tmp) %in% take_over_cols])
new_anno = tmp[order(tmp$index),new_col_order]

RG$genes = new_anno
```


## Data overview


### Samples 

```{r, results='asis', warning=FALSE, message=FALSE}
sel_cols = c("Barcode","Cy3_label","Cy5_label", "dye_swap")
tmp = ed[,sel_cols]
kable(tmp)
```

## Raw data

```{r, expr, fig.width=8, fig.height=5}
boxplot(data.frame(log2(RG$G)),main="Green", las=2)
boxplot(data.frame(log2(RG$R)),main="Red", las=2)

plotDensities(RG)
```

# Within-array normalization

```{r,  fig.width=8, fig.height=8}
MA <- normalizeWithinArrays(RG, method="loess")
MA.avg <- avereps(MA, ID=MA$genes$ProbeName)

limma::plotMA(MA,array=1,status=ifelse(MA$genes$ControlType==1,RG$genes$ProbeName,"gene"),cex=1, legend=F)
plotDensities(MA, main="Within-array normalized densities")

op <- par(mar = c(14,4,4,2) + 0.1)
boxplot(data.frame(MA$M),main="M", las=2, cex.names=0.6)
boxplot(data.frame(MA$A),main="A", las=2)
par(op)
```

```{r, cor,  fig.width=7, fig.height=5}
exp_matrix = MA.avg$M * matrix(rep(ifelse(ed$dye_swap, -1, 1),each=nrow(MA.avg$M)),nrow(MA.avg$M),ncol(MA.avg$M))

cc = cor(exp_matrix,method="spearman")
colnames(cc) = exp_design[colnames(cc), "array_label"]

pheatmap(cc)
```

```{r, echo=FALSE}
ed$Short = ed$array_label
ed$Condition = paste(ed$Cy5_label,  sep="_")
ed_orig = ed
```


## Control probes

The following control probes exist on the arrays used in this experiment:

* Corner associated (used for orientation purposes during scanning)
  * Bright corner 
  * Dark corner 
* Negative controls
  * 3xSLv1 (hairpin probe that does not hybridize well with any possible RNA)
* Positive controls
  * Human GAPDH and PGK1 probes
  * Deletion stringency probes (DCP, probe with varying number of insertions/changes with respect to reference; the number after the "_" denotes the number of differences to the reference which should correlate with lower expression)
  * E1A_r60: spike-in probes with concentrations that should cover the whole dynamic range of the array

There are a few other expression probes that are used by Agilent's feature extraction/QC pipeline. 

```{r, qc_probes,  fig.width=12, fig.height=3}
control_probes = which(RG$genes$ControlType!=0)
cp_data = RG$G[control_probes,]
cp_names = RG$genes[control_probes,]
selected_controls = ifelse(substr(cp_names$ProbeName,1,4) %in% c("ERCC", "ETG0","ETG1") | substr(cp_names$ProbeName,1,2) == "RC",F,T)

# control probes
for (i in 1:ncol(cp_data)) {
  boxplot(log2(cp_data[selected_controls,i]) ~ factor(cp_names$ProbeName[selected_controls]),las=2, main=paste("Sample",colnames(cp_data)[i]), outline=F, xlab="")
}
```

```{r}
ed=ed_orig
save(ed, RG, MA, file="../Data/Processed/Il1b_stroma_preprocessed_data.Rdata", compress = T)
```

